Schema Change: Challenge Problems

### 2024-05-11 paper.tex
harshen related work

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions. However that falls short of the range of evolutions described by \citet{ambler06} which require the addition of custom Ruby or SQL code. \citet{curino08} initiated a large thread of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17} that appears capable of handling our Extract Entity and Multiplicity Change challenges, including divergence.

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that appear capable of handling the data migration aspects of our Extract Entity and Multiplicity Change challenges. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. None of the approaches to NoSQL evolution have yet extended like their SQL cousins into updating code or rewriting queries.

Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}. Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with limited evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}. Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema evolution API~\cite{Gemstone}.

@@

>>>>

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions. However that falls short of the range of evolutions described by \citet{ambler06}, requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our Extract Entity and Multiplicity Change challenges, though Split/Merge Entity is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that may be capable of handling Extract Entity and Multiplicity Change but not Split/Merge Entity nor divergence. None of the above approaches to NoSQL evolution have yet extended into updating code or rewriting queries like their SQL cousins.

Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}. Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with some evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}. Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a complex schema evolution API~\cite{Gemstone}.

@@

Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations
